|

### 2024-04-29 paper/paper.Rmd
revise up to model arch

---
title: |
  Automated assessment of residual plots with computer vision models
type: Draft paper (Incomplete)
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
  - name: Klaus Ackermann
    affil: a
    email: Klaus.Ackermann@monash.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: ref.bib
abstract: |
  TBD.
keywords: |
  TBD
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \doublespacing
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

<!-- \newpage -->
<!-- \tableofcontents -->
<!-- \newpage -->

@@

Generating a residual plot in most statistical software is often as straightforward as executing a line of code or clicking a button. However, accurately interpreting a residual plot can be challenging. Consider Figure \ref{fig:false-finding} as an example, the residuals display a triangular shape pointing to the left. While this might suggest heteroskedasticity, it is important to avoid over-interpreting this visual pattern. In this case, the fitted regression model is correctly specified, and the triangular shape is actually a result of the skewed distribution of the predictors, rather than indicating a flaw in the model.

A residual plot can exhibit various visual features, but it is crucial to recognize that some may arise from the characteristics of predictors and the inherent randomness of the error, rather than indicating a violation of model assumptions [@li2023plot]. The concept of visual inference, as proposed by @buja2009statistical, provides an inferential framework to assess whether residual plots indeed contain visual patterns inconsistent with the model assumptions. The fundamental idea involves testing whether the actual residual plot visually differs significantly from null plots, where null plots are plotted with residuals generated from the residual rotation distribution [@langsrud2005rotation], which is a distribution consistent with the null hypothesis $H_0$ that the linear regression model is correctly specified. Typically, the visual test is accomplished through the lineup protocol, where the real residual plot is embedded within a lineup alongside several null plots. If the real residual plot can be distinguished from the lineup, it provides evidence for rejecting $H_0$.

@@

The utilization of computers to interpret data plots has a rich history, with early efforts such as "Scagnostics" by @tukey1985computer, focusing on scatter plot diagnostics. @wilkinson2005graph expanded on this work, introducing graph theoretic scagnostics, which encompassed computable measures applied to planar proximity graphs. These measures, including, but not limited to, "Outlying", "Skinny", "Stringy", "Straight", "Monotonic", "Skewed", "Clumpy", and "Striated" aimed to characterize outliers, shape, density, trend, coherence and other characteristics of the data. While this approach has been inspiring, there is a recognition [@buja2009statistical] that it may not capture all the necessary visual features that differentiate actual residual plots from null plots. A more promising alternative entails enabling machines to learn the function for extracting visual features from residual plots. Essentially, this means empowering computers to discern the crucial visual features for residual diagnostics and determining the method to extract them.

@@

## Different possible configurations of the model formula

There are various configurations of the computer vision model that can be used to assess residual plots. We discuss these configurations below based on two key components of the model formula: the input and the output format.

### Input formats

@@

Providing multiple residual plots to the model, such as a pair of plots, a triplet or a lineup is also a possible option. @chopra2005learning have shown that computer vision models designed for image comparison can assess whether a pair of images are similar or dissimilar. Applied to our specific problem, we can define null plots of a fitted regression model to be similar to each other, while considering actual residual plots to be distinct from null plots of any fitted regression model. A triplet constitutes a set of three images, denoted as $image_1$, $image_2$ and $image_3$. It is often used to predict whether $image_2$ or $image_3$ is more similar to $image_1$, proving particularly useful for establishing rankings between samples. For this setup, we can apply the same criteria to define similarity between images. However, it is important to note that these two approaches usually require additional considerations regarding the loss function and, at times, non-standard training processes due to shared weights between different convolutional blocks.

@@

We did not explore all the mentioned input formats due the considerable costs associated with data preparation and model training. Considering the implementation cost and the interpretability of the model, we settled on the single residual plot input format.

### Output formats

@@

Alternatively, the output could be a meaningful and interpretable numerical measure useful for assessing residual plots, such as the strength of suspicious visual patterns reflecting the extent of model violations and the difficulty index for identifying whether a residual plot has no issues. However, these numeric measures are often informally used in daily communication but are not typically formalized or rigorously defined. For the purpose of training a model, this numeric measure has to be quantifiable.

@@

## Distance from the good residual plots

@@

### Residual distribution

The distance measure proposed in this study takes into account the fact that we tried to measure how different a residual plot is from a good residual plot, or in other words, how different a given fitted regression model is from a correctly specified model. For the classical normal linear regression model, residuals $\boldsymbol{e}$ are derived from the fitted values $\hat{\boldsymbol{y}}$ and observed values $\boldsymbol{y}$. Suppose the data generating process is known and the regression model is correctly specified, by the Frisch-Waugh-Lowell theorem [@frisch1933partial], residuals $\boldsymbol{e}$ can also be treated as random variables and written as a linear transformation of the error $\boldsymbol{\varepsilon}$ formulated as $\boldsymbol{e} = \boldsymbol{R}\boldsymbol{\varepsilon}$, where $\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'$ is the residual operator and an idempotent matrix, $\boldsymbol{X}$ is the design matrix, $\boldsymbol{I}_n$ is a $n$ by $n$ identity matrix, and $n$ is the number of observations.

@@

In ordinary least square, the minimization of the sum of square residuals implies  $\sum_{i=1}^{n} e_i = 0$, making any residual value to be a linear combination of the remaining $n - 1$ residuals. This effectively means $\text{rank}(\boldsymbol{R}) = n - 1 < n$ and $Q$ is a degenerate multivariate distribution. To capture the characteristics of $Q$, such as moments, we can simulate a large numbers of $\boldsymbol{\varepsilon}$ and transform it to $\boldsymbol{e}$ to get the empirical estimates. For simplicity, in this study, we replaced the variance-covariance matrix of residuals $\text{cov}(\boldsymbol{e}, \boldsymbol{e}) = \boldsymbol{R}\sigma^2\boldsymbol{R}' = \boldsymbol{R}\sigma^2$ with a full-rank diagonal matrix $\text{diag}(\boldsymbol{R}\sigma^2)$, where $\text{diag}(.)$ sets the non-diagonal entries of a matrix to zeros. The resulting distribution for $\boldsymbol{Q}$ is $N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))$.

Distribution $Q$ is derived from the correctly specified model. However, if the model is misspecified, then the actual distribution of residuals denoted as $P$, will be different from $Q$. For example, if the data generating process contains variables correlated with any column of $\boldsymbol{X}$ but not included in $\boldsymbol{X}$, causing an omitted variable problem, $P$ will be different from $Q$ because the residual operator obtained from the fitted regression model will not be the same as $\boldsymbol{R}$. Besides, if the $\boldsymbol{\varepsilon}$ follows a non-normal distribution such as a multivariate log-normal distribution, $P$ will usually be skewed and has a long tail.

### Kullback-Leibler divergence of $P$ from $Q$

Define a proper distance between distributions is usually easier than define a proper distance between data plots. Given the actual residual distribution $Q$ and the reference residual distribution $P$, we used a distance measure based on Kullback-Leibler divergence [@kullback1951information] to quantify the difference between two distributions

@@

\noindent where $p(.)$ is the probability density function for distribution $P$, and $q(.)$ is the probability density function for distribution $Q$.

This distance measure was first proposed in @li2023plot. It was mainly designed for measuring the effect size of non-linearity and heteroskedasticity in a residual plot. @li2023plot have showed that, for a classical normal linear regression model that omits a necessary higher-order predictors $\boldsymbol{Z}$, and incorrectly assumes $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)$ while in fact $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})$ with $\boldsymbol{V}$ being an arbitrary symmetric positive semi-definite matrix, $Q$ can be represented as $N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}))$. Note that the variance-covariance matrix is replaced with the diagonal matrix to ensure it is a full-rank matrix.

@@

\begin{align}
\label{eq:kl-2}
D_{KL} &= \frac{1}{2}\left(\log\frac{|\text{diag}(\boldsymbol{W})|}{|\text{diag}(\boldsymbol{R}\sigma^2)|} - n + \text{tr}(\text{diag}(\boldsymbol{W})^{-1}\text{diag}(\boldsymbol{R}\sigma^2)) + \boldsymbol{\mu}_z'(\text{diag}(\boldsymbol{W}))^{-1}\boldsymbol{\mu}_z\right),
\end{align}

\noindent where $\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z$, and $\boldsymbol{W} = \boldsymbol{R}\boldsymbol{V}\boldsymbol{R}$. The assumed error variance $\sigma^2$ is set to be $\text{tr}(\boldsymbol{V})/n$, which is the expectation of the estimated variance.

### Evaluation of Kullback-Leibler divergence for non-normal $P$

For non-normal error $\boldsymbol{\varepsilon}$, the actual residual distribution $P$ is unlikely to be a multivariate normal distribution. Thus, equation \ref{eq:kl-2} given in @li2023plot will not be applicable to models violating the normality assumption.

@@

### Approximation of the distance measure

In the previous sections, we have defined a distance measure given in equation \ref{eq:kl-0} for quantifying the difference between the actual residual distribution $P$ and an ideal reference distribution $Q$. You may have noticed that this distance measure can only be computed when the data generating process is known. In reality, we often have no knowledge about the data generating process, otherwise we do not need to fit a linear regression model in the first place.

We tried to train a computer vision model to approximate this distance measure with a residual plot. Let $D$ be the result of equation \ref{eq:kl-0}, and our estimator $\hat{D}$ is formulated as

@@

\noindent where $V_{h \times w}(.)$ is a plotting function that saves a residuals vs fitted values plot with fixed aesthetic as an image with $h \times w$ pixels and three colour channels, $f_{CV}(.)$ is a computer vision model which takes an $h \times w$ image as input and predicts the distance in the domain $[0, +\infty)$.

With the approximated distance $\hat{D}$, we will be able to know how different the underlying distribution of the residuals is from a good residual distribution. The approximated distance $\hat{D}$ can also be used as an index of the model violations. It also provides information for the strength of the visual signal embedded in the residual plot.

The approximated distance $\hat{D}$ is not expected to be the same as the original distance $D$. This is largely because information contained in a single residual plot is limited and it may not be able to summarise all the important characteristics of the residual distribution. For a given residual distribution $P$, we can generate many different residual plots. Some of them share similar visual patterns, but some of them could be visually very different from the rest, especially for regression models with small $n$. This suggests the error of the approximation will vary depends on whether the observed residual plot is representative or not.

### Model Violations Index (MVI) {#sec:model-violations-index}

In Section \ref{output-formats}, we discuss the output formats of the computer vision model presented in this paper. One notable advantage highlighted is its interpretability. The approximated distance $\hat{D}$ is an estimator of the difference between the actual residual distribution and the reference residual distribution. This difference primarily arises from deviations in model assumptions. The magnitude of $D$ directly reflects the degree of these deviations, thus making $\hat{D}$ instrumental in forming a model violation index.

Note that the Kullback-Leibler divergence might be influenced by the number of random variables involved in its evaluation. Generally, a larger number of observations will lead to a greater distance $D$. However, this does not imply that $\hat{D}$ fails to accurately represent the extent of model violations. In fact, when examining residual plots with more observations, we often observe a stronger visual signal strength, as the underlying patterns are more likely to be revealed, except in cases of significant overlapping.

@@

## Statistical testing based on the approximated distance

### Lineup evaluation

@@

To address this issue while adhering to the principle of visual inference, we can compare the estimated distance $\hat{D}$ to the approximated distances for the null plots in a lineup. Specifically, if a lineup comprises 20 plots, the null hypothesis $H_0$ will be rejected if $\hat{D}$ exceeds the maximum approximated distance among the $m - 1$ null plots, denoted as $\max\limits_{1 \leq i \leq m-1} {\hat{D}_{null}^{(i)}}$, where $\hat{D}_{null}^{(i)}$ represents the approximated distance for the $i$-th null plot. This approach is equivalent to the typical lineup protocol requiring a 95% significance level, where $H_0$ is rejected if the data plot is identified as the most distinct plot by the sole participant. The approximated distance serves as a metric to quantify the difference between the data plot and the null plots, as intended.

For lineups consisting of more than 20 plots, the 95% significance level can be maintained if the number of plots is a multiple of 20. Specifically, for lineups comprising $20p$ plots, where $p$ is a positive integer, we reject $H_0$ if $\hat{D}$ exceeds 95% ${\hat{D}_{null}^{(i)}}$ for $i = 1, \ldots, 20p-1$. The p-value in this case is given by $\frac{1}{m} + \frac{1}{m}\sum{i=1}^{m-1} I\left(\hat{D}_{null}^{(i)} > \hat{D}\right)$, where $I(\cdot)$ is the indicator function.

Moreover, if the number of plots in a lineup, denoted by $m$, is sufficiently large, the empirical distribution of ${\hat{D}_{null}^{(i)}}$ can be viewed as an approximation of the null distribution of the approximated distance. Consequently, quantiles of the null distribution can be estimated using the sample quantiles available in statistical software such as R, and these quantiles can be utilized for decision-making purposes. The details of the sample quantile computation can be found in @hyndman1996sample. For instance, if $\hat{D}$ is greater than or equal to the 95% sample quantile, denoted as $Q_{null}(0.95)$, we can conclude that the approximated distance for the actual residual plot is significantly different from the approximated distance for null plots with a 95% significance level. Based on our experience, to obtain a stable estimate of the 95% quantile, the number of null plots, $n_{null}$, typically needs to be at least 100. However, if the null distribution exhibits a long tail, a larger number of null plots may be required. Alternatively, a p-value can be used to represent the probability of observing an event equally or more extreme than the given event under the null hypothesis $H_0$, and it can be estimated by $\frac{1}{m}\sum_{i=1}^{m-1}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)$.

@@

### Bootstrapping the approximated distance

@@

Similarly, bootstrap can be applied on the approximated distance $\hat{D}$. For each refitted model $M_{boot}^{(i)}$, there will be an associated residual plot $V_{boot}^{(i)}$ which can be fed into the computer vision model to obtain $\hat{D}_{boot}^{(i)}$, where $i = 1,...,n_{boot}$, and $n_{boot}$ is the number of bootstrapped samples. If we are interested in the variation of $\hat{D}$, we can use $\hat{D}_{boot}^{(i)}$ to estimate a confidence interval.

@@

## Generation of training data

### Data generating process

@@

### Computation of scagnostics

@@

For each generated residual plot, we computed four scagnostics – "Monotonic," "Sparse," "Splines," and "Striped" – using the `cassowaryr` R package [@mason2022cassowaryr]. These computed measures, along with the number of observations from the fitted model, were provided as the second input for the computer vision model. Although other scagnostics are informative, they are currently unavailable due to a fatal bug in the compiled C program of the `interp` R package [@Albrecht2023interp], which may unpredictably crash the process. For reproducibility, we excluded these scagnostics from the training data.

### Crafting a balanced training set

@@

## Architecture of the computer vision model

@@

## Training and hyperparameter tuning

@@

## Model evaluation methods

@@

## Optimized hyperparameters

@@

## Demostration of MVI

As described in Section \ref{sec:model-violations-index} a Model Violations Index (MVI) can be constructed using $\hat{D}$. This section provides guidelines and examples of MVI.

@@

In order to check the validity of the proposed computer vision model, residual plots presented in the human subject experiment conducted by @li2023plot will be assessed. This study has collected $7974$ human responses to $1152$ lineups. Each lineup contains one randomly placed actual residual plot and 19 null plots. Among the $1152$ lineups, $24$ are attention check lineups in which the visual patterns are designed to be extremely obvious and very different from the corresponding to null plots, $36$ are null lineups where all the lineups consist of only null plots, $279$ are lineups with uniform predictor distribution evaluated by $11$ participants, and the remaining $813$ are lineups with discrete, skewed or normal predictor distribution evaluated by $5$ participants. Attention check lineups and null lineups will not be assessed in the following analysis.

@@

For each lineup used in @li2023plot, there is one actual residual plot and 19 null plots. While the distance $D$ for the actual residual plot depends on the underlying data generating process, the distance $D$ for the null plots is zero. We have used our optimized computer vision model to predict the approximated distance for both the actual residual plots and the null plots. To have a fair comparison, $H_0$ will be rejected if the actual residual plot has the greatest approximated distance among all plots in a lineup. Additionally, the appropriate conventional tests including the Ramsey Regression Equation Specification Error Test (RESET) [@ramsey1969tests] for non-linearity and the Breusch-Pagan test  [@breusch1979simple] for heteroskedasticity were applied on the same data for comparison.

The performance metrics of $\hat{D}$ for actual residual plots are outlined in Table \ref{tab:experiment-performance}. It's notable that all performance metrics are slightly worse than those evaluated on the test data. Nevertheless, the mean absolute error remains at a low level, and the linear correlation between the prediction and the true value remains very high. Lineups with non-linearity issues are more challenging to predict than those with heteroskedasticity issues.

>>>>

---
title: |
  Automated assessment of residual plots with computer vision models
type: Draft paper (Incomplete)
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
  - name: Klaus Ackermann
    affil: a
    email: Klaus.Ackermann@monash.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: ref.bib
abstract: |
  TBD.
keywords: |
  TBD
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

@@

Generating a residual plot in most statistical software is often as straightforward as executing a line of code or clicking a button. However, accurately interpreting a residual plot can be challenging. A residual plot can exhibit various visual features, but it is crucial to recognize that some may arise from the characteristics of predictors and the inherent randomness of the error, rather than indicating a violation of model assumptions [@li2023plot]. Consider Figure \ref{fig:false-finding} as an example, the residuals display a triangular shape pointing to the left. While this might suggest heteroskedasticity, it is important to avoid over-interpreting this visual pattern. In this case, the fitted regression model is correctly specified, and the triangular shape is actually a result of the skewed distribution of the predictors, rather than indicating a flaw in the model.

The concept of visual inference, as proposed by @buja2009statistical, provides an inferential framework to assess whether residual plots indeed contain visual patterns inconsistent with the model assumptions. The fundamental idea involves testing whether the true residual plot visually differs significantly from null plots, where null plots are plotted with residuals generated from the residual rotation distribution [@langsrud2005rotation], which is a distribution consistent with the null hypothesis $H_0$ that the linear regression model is correctly specified. Typically, the visual test is accomplished through the lineup protocol, where the real residual plot is embedded within a lineup alongside several null plots. If the real residual plot can be distinguished from the lineup, it provides evidence for rejecting $H_0$.

@@

The utilization of computers to interpret data plots has a rich history, with early efforts such as "Scagnostics" by @tukey1985computer, focusing on scatter plot diagnostics. @wilkinson2005graph expanded on this work, introducing graph theoretic scagnostics, which encompassed computable measures applied to planar proximity graphs. These measures, including, but not limited to, "Outlying", "Skinny", "Stringy", "Straight", "Monotonic", "Skewed", "Clumpy", and "Striated" aimed to characterize outliers, shape, density, trend, coherence and other characteristics of the data. While this approach has been inspiring, there is a recognition [@buja2009statistical] that it may not capture all the necessary visual features that differentiate true residual plots from null plots. A more promising alternative entails enabling machines to learn the function for extracting visual features from residual plots. Essentially, this means empowering computers to discern the crucial visual features for residual diagnostics and determining the method to extract them.

@@

## Model Formulas

There are various configurations of the computer vision model that can be used to assess residual plots. We discuss these configurations below focusing on two key components of the model formula: the input and the output format.

### Input Formats

@@

Providing multiple residual plots to the model, such as a pair of plots, a triplet or a lineup is also a possible option. @chopra2005learning have shown that computer vision models designed for image comparison can assess whether a pair of images are similar or dissimilar. Applied to our specific problem, we can define null plots of a fitted regression model to be similar to each other, while considering true residual plots to be distinct from null plots of any fitted regression model. A triplet constitutes a set of three images, denoted as $image_1$, $image_2$ and $image_3$. It is often used to predict whether $image_2$ or $image_3$ is more similar to $image_1$, proving particularly useful for establishing rankings between samples. For this setup, we can apply the same criteria to define similarity between images. However, it is important to note that these two approaches usually require additional considerations regarding the loss function and, at times, non-standard training processes due to shared weights between different convolutional blocks.

@@

We did not explore all the mentioned input formats due to the significant costs tied to data preparation and model training. Taking into account the implementation cost and the need for model interpretability, we settled on the single residual plot input format.

### Output Formats

@@

Alternatively, the output could be a meaningful and interpretable numerical measure useful for assessing residual plots, such as the strength of suspicious visual patterns reflecting the extent of model violations, or the difficulty index for identifying whether a residual plot has no issues. However, these numeric measures are often informally used in daily communication but are not typically formalized or rigorously defined. For the purpose of training a model, this numeric measure has to be quantifiable.

@@

## Distance Between Residual Plots

@@

### Residual Distribution

The distance measure proposed in this study takes into account the fact that we tried to measure how different a residual plot is from a good residual plot, or in other words, how different a given fitted regression model is from a correctly specified model.

For a classical normal linear regression model, residuals $\boldsymbol{e}$ are derived from the fitted values $\hat{\boldsymbol{y}}$ and observed values $\boldsymbol{y}$. Suppose the data generating process is known and the regression model is correctly specified, by the Frisch-Waugh-Lowell theorem [@frisch1933partial], residuals $\boldsymbol{e}$ can also be treated as random variables and written as a linear transformation of the error $\boldsymbol{\varepsilon}$ formulated as $\boldsymbol{e} = \boldsymbol{R}\boldsymbol{\varepsilon}$, where $\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'$ is the residual operator and an idempotent matrix, $\boldsymbol{X}$ is the design matrix, $\boldsymbol{I}_n$ is a $n$ by $n$ identity matrix, and $n$ is the number of observations.

@@

Suppose the design matrix $\boldsymbol{X}$ has linearly independent columns, the trace of the hat matrix will equal to the number of columns in $\boldsymbol{X}$ denoted as $k$. As a result, the rank of $\boldsymbol{R}$ is $n - k$, and $Q$ is a degenerate multivariate distribution. To capture the characteristics of $Q$, such as moments, we can simulate a large numbers of $\boldsymbol{\varepsilon}$ and transform it to $\boldsymbol{e}$ to get the empirical estimates. For simplicity, in this study, we replaced the variance-covariance matrix of residuals $\text{cov}(\boldsymbol{e}, \boldsymbol{e}) = \boldsymbol{R}\sigma^2\boldsymbol{R}' = \boldsymbol{R}\sigma^2$ with a full-rank diagonal matrix $\text{diag}(\boldsymbol{R}\sigma^2)$, where $\text{diag}(.)$ sets the non-diagonal entries of a matrix to zeros. The resulting distribution for $\boldsymbol{Q}$ is $N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))$.

Distribution $Q$ is derived from the correctly specified model. However, if the model is misspecified, then the actual distribution of residuals denoted as $P$, will be different from $Q$. For example, if the data generating process contains variables correlated with any column of $\boldsymbol{X}$ but missing from $\boldsymbol{X}$, causing an omitted variable problem, $P$ will be different from $Q$ because the residual operator obtained from the fitted regression model will not be the same as $\boldsymbol{R}$. Besides, if the $\boldsymbol{\varepsilon}$ follows a non-normal distribution such as a multivariate log-normal distribution, $P$ will usually be skewed and has a long tail.

### Distance of $P$ from $Q$

Define a proper distance between distributions is usually easier than define a proper distance between data plots. Given the true residual distribution $Q$ and the reference residual distribution $P$, we used a distance measure based on Kullback-Leibler divergence [@kullback1951information] to quantify the difference between two distributions

@@

\noindent where $p(.)$ and $q(.)$ are the probability density functions for distribution $P$ and distribution $Q$ respectively.

This distance measure was first proposed in @li2023plot. It was mainly designed for measuring the effect size of non-linearity and heteroskedasticity in a residual plot. @li2023plot have derived that, for a classical normal linear regression model that omits a necessary higher-order predictors $\boldsymbol{Z}$, and incorrectly assumes $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)$ while in fact $\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})$ with $\boldsymbol{V}$ being an arbitrary symmetric positive semi-definite matrix, $Q$ can be represented as $N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}))$. Note that the variance-covariance matrix is replaced with the diagonal matrix to ensure it is a full-rank matrix.

@@

\begin{align}
\label{eq:kl-2}
D_{KL} &= \frac{1}{2}\left(\log\frac{|\text{diag}(\boldsymbol{W})|}{|\text{diag}(\boldsymbol{R}\sigma^2)|} - n + tr(\text{diag}(\boldsymbol{W})^{-1}\text{diag}(\boldsymbol{R}\sigma^2)) + \boldsymbol{\mu}_z'(\text{diag}(\boldsymbol{W}))^{-1}\boldsymbol{\mu}_z\right),
\end{align}

\noindent where $\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z$, and $\boldsymbol{W} = \boldsymbol{R}\boldsymbol{V}\boldsymbol{R}$. The assumed error variance $\sigma^2$ is set to be $tr(\boldsymbol{V})/n$, which is the expectation of the estimated variance.

### Non-normal $P$

For non-normal error $\boldsymbol{\varepsilon}$, the true residual distribution $P$ is unlikely to be a multivariate normal distribution. Thus, equation \ref{eq:kl-2} given in @li2023plot will not be applicable to models violating the normality assumption.

@@

## Distance Estimation

In the previous sections, we have defined a distance measure given in equation \ref{eq:kl-0} for quantifying the difference between the true residual distribution $P$ and an ideal reference distribution $Q$. You may have noticed that this distance measure can only be computed when the data generating process is known. In reality, we often have no knowledge about the data generating process, otherwise we do not need to fit a linear regression model in the first place.

We tried to train a computer vision model to estimate this distance measure with a residual plot. Let $D$ be the result of equation \ref{eq:kl-0}, and our estimator $\hat{D}$ is formulated as

@@

\noindent where $V_{h \times w}(.)$ is a plotting function that saves a residuals vs fitted values plot with fixed aesthetic as an RGB image with $h \times w$ pixels, $f_{CV}(.)$ is a computer vision model which takes an $h \times w$ image as input and predicts the distance in the domain $[0, +\infty)$.

With the estimated distance $\hat{D}$, we will be able to know how different the underlying distribution of the residuals is from a good residual distribution. $\hat{D}$ can also be used as an index of the model violations indicating the strength of the visual signal embedded in the residual plot.

It is not expected that $\hat{D}$ will be equal to original distance $D$. This is largely because information contained in a single residual plot is limited and it may not be able to summarise all the important characteristics of the residual distribution. For a given residual distribution $P$, many different residual plots can be simulated, where many will share similar visual patterns, but some of them could be visually very different from the rest, especially for regression models with small $n$. This suggests the error of the estimation will vary depends on whether the input residual plot is representative or not.

## Model Violations Index (MVI) {#sec:model-violations-index}

$\hat{D}$ is an estimator of the difference between the true residual distribution and the reference residual distribution. This difference primarily arises from deviations in model assumptions. The magnitude of $D$ directly reflects the degree of these deviations, thus making $\hat{D}$ instrumental in forming a model violations index.

Note that Equation \ref{eq:kl-0} might be influenced by the number of random variables involved in its evaluation. Generally, a larger number of observations will lead to a greater distance $D$. However, this does not imply that $\hat{D}$ fails to accurately represent the extent of model violations. In fact, when examining residual plots with more observations, we often observe a stronger visual signal strength, as the underlying patterns are more likely to be revealed, except in cases of significant overlapping.

@@

## Statistical testing

### Lineup Evaluation

@@

To address this issue while adhering to the principle of visual inference, we can compare the estimated distance $\hat{D}$ to the estimated distances for the null plots in a lineup. Specifically, if a lineup comprises 20 plots, the null hypothesis $H_0$ will be rejected if $\hat{D}$ exceeds the maximum estimated distance among the $m - 1$ null plots, denoted as $\max\limits_{1 \leq i \leq m-1} {\hat{D}_{null}^{(i)}}$, where $\hat{D}_{null}^{(i)}$ represents the estimated distance for the $i$-th null plot. This approach is equivalent to the typical lineup protocol requiring a 95% significance level, where $H_0$ is rejected if the data plot is identified as the most distinct plot by the sole participant. The estimated distance serves as a metric to quantify the difference between the data plot and the null plots, as intended.

For lineups consisting of more than 20 plots, the 95% significance level can be maintained if the number of plots is a multiple of 20. Specifically, for lineups comprising $20t$ plots, where $l$ is a positive integer, we reject $H_0$ if $\hat{D}$ exceeds 95% ${\hat{D}_{null}^{(i)}}$ for $i = 1, \ldots, 20t-1$. The $p$-value in this case is given by $\frac{1}{20t} + \frac{1}{20t}\sum{i=1}^{20t-1} I\left(\hat{D}_{null}^{(i)} > \hat{D}\right)$, where $I(\cdot)$ is the indicator function.

Moreover, if the number of plots in a lineup, denoted by $m$, is sufficiently large, the empirical distribution of ${\hat{D}_{null}^{(i)}}$ can be viewed as an approximation of the null distribution of the approximated distance. Consequently, quantiles of the null distribution can be estimated using the sample quantiles available in statistical software such as R, and these quantiles can be utilized for decision-making purposes. The details of the sample quantile computation can be found in @hyndman1996sample. For instance, if $\hat{D}$ is greater than or equal to the 95% sample quantile, denoted as $Q_{null}(0.95)$, we can conclude that the estimated distance for the true residual plot is significantly different from the estimated distance for null plots with a 95% significance level. Based on our experience, to obtain a stable estimate of the 95% quantile, the number of null plots, $n_{null}$, typically needs to be at least 100. However, if the null distribution exhibits a long tail, a larger number of null plots may be required. Alternatively, a $p$-value can be used to represent the probability of observing an event equally or more extreme than the given event under the null hypothesis $H_0$, and it can be estimated by $\frac{1}{m}\sum_{i=1}^{m-1}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)$.

@@

### Bootstrapping

@@

Similarly, bootstrap can be applied on the estimated distance $\hat{D}$. For each refitted model $M_{boot}^{(i)}$, there will be an associated residual plot $V_{boot}^{(i)}$ which can be fed into the computer vision model to obtain $\hat{D}_{boot}^{(i)}$, where $i = 1,...,n_{boot}$, and $n_{boot}$ is the number of bootstrapped samples. If we are interested in the variation of $\hat{D}$, we can use $\hat{D}_{boot}^{(i)}$ to estimate a confidence interval.

@@

## Data Genearation

### Simulation Scheme

@@

### Scagnostics

@@

For each generated residual plot, we computed four scagnostics – "Monotonic," "Sparse," "Splines," and "Striped" – using the `cassowaryr` R package [@mason2022cassowaryr]. These computed measures, along with the number of observations from the fitted model, were provided as the second input for the computer vision model. While other scagnostics provide valuable insights, they come with high computational costs and are not suitable for quick inference.

### Balanced Dataset

@@

## Model Architecture

@@

## Model Training

@@

## Evaluation Methods

@@

understand what the convolutional layer is actually doing (read vgg16 doc)

@@

@@

In order to check the validity of the proposed computer vision model, residual plots presented in the human subject experiment conducted by @li2023plot will be assessed. This study has collected $7974$ human responses to $1152$ lineups. Each lineup contains one randomly placed true residual plot and 19 null plots. Among the $1152$ lineups, $24$ are attention check lineups in which the visual patterns are designed to be extremely obvious and very different from the corresponding to null plots, $36$ are null lineups where all the lineups consist of only null plots, $279$ are lineups with uniform predictor distribution evaluated by $11$ participants, and the remaining $813$ are lineups with discrete, skewed or normal predictor distribution evaluated by $5$ participants. Attention check lineups and null lineups will not be assessed in the following analysis.

@@

For each lineup used in @li2023plot, there is one true residual plot and 19 null plots. While the distance $D$ for the true residual plot depends on the underlying data generating process, the distance $D$ for the null plots is zero. We have used our optimized computer vision model to predict the approximated distance for both the true residual plots and the null plots. To have a fair comparison, $H_0$ will be rejected if the true residual plot has the greatest approximated distance among all plots in a lineup. Additionally, the appropriate conventional tests including the Ramsey Regression Equation Specification Error Test (RESET) [@ramsey1969tests] for non-linearity and the Breusch-Pagan test  [@breusch1979simple] for heteroskedasticity were applied on the same data for comparison.

The performance metrics of $\hat{D}$ for true residual plots are outlined in Table \ref{tab:experiment-performance}. It's notable that all performance metrics are slightly worse than those evaluated on the test data. Nevertheless, the mean absolute error remains at a low level, and the linear correlation between the prediction and the true value remains very high. Lineups with non-linearity issues are more challenging to predict than those with heteroskedasticity issues.
